\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{array}
\usepackage{url}
\usepackage{hyperref}

\hypersetup{
    colorlinks=true,
    linkcolor=black,
    citecolor=black,
    urlcolor=black
}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Enterprise VPN Technologies: A Comparative Survey of GRE, MPLS, and IPSec Solutions}

\author{
    \IEEEauthorblockN{[Author 1 Name]}
    \IEEEauthorblockA{\textit{Department of Computer Science} \\
    \textit{Habib University}\\
    Karachi, Pakistan \\
    student1@stu.habib.edu.pk}
    \and
    \IEEEauthorblockN{[Author 2 Name]}
    \IEEEauthorblockA{\textit{Department of Computer Science} \\
    \textit{Habib University}\\
    Karachi, Pakistan \\
    student2@stu.habib.edu.pk}
    \and
    \IEEEauthorblockN{[Author 3 Name]}
    \IEEEauthorblockA{\textit{Department of Computer Science} \\
    \textit{Habib University}\\
    Karachi, Pakistan \\
    student3@stu.habib.edu.pk}
}

\maketitle

% ---------------------------------------------------------------
\begin{abstract}
Virtual Private Networks (VPNs) have become a foundational component
of enterprise communication infrastructure, providing secure and
reliable connectivity across geographically distributed sites. This
paper presents a comprehensive literature review of three widely
deployed enterprise VPN technologies: Generic Routing Encapsulation
(GRE), Multi-Protocol Label Switching (MPLS), and Internet Protocol
Security (IPSec). We examine the architectural principles, security
properties, and deployment characteristics of each technology, along
with hybrid combinations such as GRE over IPSec. By analyzing recent
research and practical implementations, we identify the strengths and
limitations of each approach and provide a comparative discussion to
guide enterprises in selecting the most appropriate VPN solution for
their specific requirements. The findings suggest that no single
technology is universally superior; rather, the choice depends on
factors including security requirements, scalability needs, quality
of service demands, and cost constraints.
\end{abstract}

\begin{IEEEkeywords}
VPN, GRE, IPSec, MPLS, enterprise network, tunneling, network security, L3VPN
\end{IEEEkeywords}

% ---------------------------------------------------------------
\section{Introduction}

In the modern enterprise environment, organizations are no longer
confined to a single physical location. Branch offices, remote
workers, and partner organizations span cities, countries, and even
continents. This geographic dispersion creates a fundamental
challenge: how do we connect these sites securely and efficiently
over public network infrastructure, without exposing sensitive data
to unauthorized parties?

Virtual Private Networks offer a practical answer to this challenge.
A VPN uses a shared or public infrastructure, such as the Internet,
to create a private communication channel that mimics the behavior
of a dedicated physical network~\cite{rosu2012}. From the perspective
of the end user, traffic appears to flow through a private pipe, even
though it physically travels through shared infrastructure. VPNs
provide the confidentiality, integrity, and availability properties
that enterprise operations demand.

Three technologies dominate enterprise VPN deployments today. The
first is Generic Routing Encapsulation (GRE), a tunneling protocol
defined in RFC~2784~\cite{rfc2784} that encapsulates one network
protocol inside another. GRE is lightweight and flexible but
provides no native encryption. The second is IPSec, a comprehensive
security framework defined in RFC~4301~\cite{rfc4301} that provides
authentication and encryption at the network layer. IPSec is the de
facto standard for securing VPN traffic over the Internet. The third
is MPLS, defined in RFC~3031~\cite{rfc3031}, a high-performance
packet forwarding mechanism that enables service providers to deliver
scalable, multi-tenant VPN services with strong traffic isolation
and Quality of Service (QoS) guarantees.

Each technology has found a distinct place in enterprise networking.
GRE is commonly used for its simplicity and routing flexibility.
IPSec is deployed wherever strong encryption is required. MPLS is
the preferred solution for service providers and large enterprises
that need scalability and traffic engineering. In many real-world
deployments, these technologies are combined --- most notably in the
form of GRE over IPSec --- to leverage the strengths of more than
one approach simultaneously.

This paper surveys the literature on all three technologies and their
combinations. The goal is to provide a clear, structured understanding
of how each technology works, what it is good for, and where it falls
short. The rest of the paper is organized in such a way that
Section~\ref{sec:relatedwork} discusses related work found in the
literature, followed by a discussion in Section~\ref{sec:discussion}
and the paper is concluded in Section~\ref{sec:conclusion}.

% ---------------------------------------------------------------
\section{Background}

Before reviewing specific technologies, it is useful to briefly
establish the general VPN taxonomy that the literature commonly uses.
VPNs can be classified along several dimensions. By responsibility,
they are either customer-provisioned (the customer manages the VPN
endpoints) or provider-provisioned (the service provider manages the
VPN infrastructure)~\cite{akinsanya2024}. By layer of operation, they
are either Layer-2 VPNs, which operate at the data-link layer and
present a virtual switch to the customer, or Layer-3 VPNs, which
operate at the network layer and involve the service provider in IP
routing~\cite{fayed2024}. By access model, they are either
site-to-site VPNs, which connect fixed locations, or remote access
VPNs, which serve mobile and home users~\cite{rosu2012}.

The three technologies covered in this paper map to different
positions in this taxonomy. GRE is a customer-side tunneling
mechanism that operates at Layer-3. IPSec is a security framework
that operates at Layer-3 and is used in both customer and provider
deployments. MPLS is primarily a provider-side technology that
supports both Layer-2 and Layer-3 VPNs at significant scale.

% ---------------------------------------------------------------
\section{Related Work}
\label{sec:relatedwork}

\subsection{Tunnel-Based VPN Technologies}

\subsubsection{Generic Routing Encapsulation (GRE)}

GRE is one of the oldest and most widely used tunneling protocols
in enterprise networking. It was standardized by Cisco and later
published by the IETF in RFC~2784~\cite{rfc2784}. The fundamental
idea behind GRE is simple: a packet of one protocol, such as IPv6,
is encapsulated inside a GRE header and then placed inside an outer
IPv4 packet for delivery across a network that would not otherwise
support the inner protocol. At the receiving end, the outer header
and GRE header are removed and the original packet is forwarded
normally.

GRE tunnels are stateless, meaning the tunnel itself does not
maintain session information. This makes GRE lightweight and easy
to configure. A common real-world use case, demonstrated in the
work of Fahriza et al.~\cite{fahriza2026}, is IPv6-over-IPv4
tunneling, where an organization needs to deploy IPv6 across
branches but the WAN infrastructure only supports IPv4. GRE handles
this encapsulation cleanly and enables OSPF routing to run
end-to-end across the tunnel, ensuring efficient convergence.

However, GRE has one significant limitation: it provides no
encryption and no authentication on its own. A GRE tunnel passes
traffic as plaintext, meaning any attacker who can observe traffic
on the underlying network can read the contents of the tunnel. GRE
was never intended as a security mechanism; it is a routing and
encapsulation tool.

Uddin et al.~\cite{uddin2021} conducted an analysis of GRE over
IPv6 networks and confirmed that while GRE provides the routing
flexibility needed to connect heterogeneous networks, the absence
of native security makes it unsuitable as a standalone enterprise
VPN solution where data confidentiality is required. Their work
reinforces the view that GRE is best understood as a building block
to be combined with security protocols, rather than a complete VPN
solution in itself.

Ogudo~\cite{ogudo2019} examined GRE and IPSec tunneling protocols
specifically for secure communication over public networks and
concluded that GRE alone is appropriate only in trusted or
internally controlled environments. For public Internet traversal,
the combination of GRE with IPSec is necessary.

\subsubsection{IPSec Protocol Suite}

IPSec is a comprehensive framework defined across multiple RFCs,
with the architecture specified in RFC~4301~\cite{rfc4301}. Unlike
GRE, IPSec was designed from the ground up as a security solution.
It operates at the IP layer and is transparent to applications
above it, which means no application modification is required for
IPSec to protect traffic.

IPSec provides three core security services. The first is
confidentiality, achieved through encryption using algorithms such
as AES. The second is data integrity, achieved through keyed hash
functions that verify packets have not been tampered with in
transit. The third is authentication, which verifies the identity
of communicating parties. These services are delivered through two
protocols: the Authentication Header (AH), which provides integrity
and authentication but does not encrypt the payload, and the
Encapsulating Security Payload (ESP), which provides all three
services including confidentiality, making it the more commonly
deployed option in practice.

IPSec operates in two modes. In transport mode, only the payload
of the IP packet is protected and the original IP header is
preserved. In tunnel mode, the entire original IP packet is
encapsulated inside a new IP packet and fully protected. Tunnel
mode is the standard choice for gateway-to-gateway VPNs, where
the VPN endpoints are different from the actual source and
destination hosts~\cite{rosu2012}.

Key management in IPSec is handled by the Internet Key Exchange
(IKE) protocol. IKE version 1 was widely deployed but has known
vulnerabilities. IKEv2 resolves these issues and offers improved
authentication and negotiation efficiency. Most modern deployments
use IKEv2.

Nair~\cite{nair2024} provided a detailed examination of IPSec
integration for data security and concluded that IPSec with ESP in
tunnel mode is the most robust option for enterprise site-to-site
VPNs. The combination of AES-256 for encryption and SHA-256 for
integrity provides strong protection against both passive
eavesdropping and active tampering attacks.

Rosu et al.~\cite{rosu2012} studied a large geographically
dispersed enterprise and implemented an IPSec-based VPN architecture
connecting a headquarters in Bucharest with hundreds of branch
offices across Romania. Their implementation used IPSec tunnel mode
with DES and 3DES encryption deployed on Cisco routers. The authors
found that IPSec met the confidentiality and integrity requirements
of the organization and integrated cleanly with the existing routing
infrastructure. The deployment also demonstrated that open-source
network monitoring tools like Nagios can effectively manage the
health of large-scale IPSec VPN networks.

Sankaran~\cite{sankaran2023} examined cross-protocol attacks on
IPSec-based VPNs and identified several attack vectors including
IKEv1 downgrade attacks and traffic analysis. This work highlights
that while the IPSec cryptographic core is strong, the overall
security depends on careful configuration, particularly in the
choice of IKE version and cipher suites. The author recommends
mandatory use of IKEv2 and avoidance of legacy cipher suites such
as 3DES in any new deployment.

\subsubsection{GRE over IPSec (Hybrid Approach)}

Because GRE lacks security and IPSec lacks routing flexibility,
combining the two has become a standard practice in enterprise
networking. In a GRE-over-IPSec configuration, a GRE tunnel is
first established between two routers, and then IPSec is applied
to encrypt all traffic flowing through that tunnel. The result is
a tunnel that supports dynamic routing protocols like OSPF and
EIGRP --- which cannot run directly over a raw IPSec tunnel ---
while also providing full encryption of all encapsulated traffic.

Fahriza et al.~\cite{fahriza2026} implemented exactly this
architecture for a private network simulation in Indonesia. Their
topology consisted of three regional branch offices connected
through GRE tunnels with IPSec encryption, using OSPF for routing
and ISAKMP for IPSec key management. Network performance was
evaluated by measuring QoS metrics including delay, jitter, and
throughput. The simulation results showed that the GRE-IPSec
combination successfully established secure connections between
all branch sites, and network quality remained acceptable even
when tunneling was active. The authors also conducted a
techno-economic analysis and found the implementation financially
viable, with an Internal Rate of Return of 84.80\% and a payback
period of approximately 11 months.

Kumar et al.~\cite{kumar2020} also studied IPSec with encapsulation
and found that the combination of GRE and IPSec, when properly
configured, provides an effective balance between routing
functionality and security. Their work specifically examined how
multicast and broadcast traffic, which IPSec cannot carry
directly, can be supported through GRE encapsulation before
IPSec encryption.

The main tradeoff in GRE-over-IPSec is overhead. Each packet
carries both a GRE header (4 bytes minimum) and IPSec headers.
This additional overhead reduces the effective MTU of the tunnel
and increases CPU load on the router. For environments with high
throughput requirements or latency-sensitive applications, this
overhead must be accounted for in capacity planning.

\subsection{Provider-Managed VPN Technologies: MPLS}

\subsubsection{MPLS Architecture and Core Concepts}

Multi-Protocol Label Switching was developed by the IETF to address
the limitations of traditional hop-by-hop IP routing. MPLS is
defined in RFC~3031~\cite{rfc3031} and introduces a 32-bit label
inserted between the Layer-2 header and the Layer-3 IP header.
Routers in an MPLS domain --- called Label Switching Routers (LSRs)
--- forward packets based on this label rather than performing a
full IP routing table lookup at each hop. This label-based
forwarding is faster and more predictable than conventional IP
routing.

The MPLS label structure consists of a 20-bit label value, 3~bits
for experimental (EXP) use, 1~bit for bottom-of-stack indication,
and 8~bits for Time-to-Live (TTL). Labels are assigned and
distributed among routers using the Label Distribution Protocol
(LDP) or the Resource Reservation Protocol with Traffic Engineering
extensions (RSVP-TE).

At the edges of the MPLS domain, Provider Edge (PE) routers
receive traffic from Customer Edge (CE) routers and assign labels
before forwarding into the MPLS core. Provider (P) routers in the
core only perform label switching and have no knowledge of customer
VPN information. At the far edge, the receiving PE router removes
the label and delivers traffic to the destination CE router. This
penultimate hop popping mechanism, where the label is removed one
hop before the destination, further reduces forwarding overhead.

Mungse~\cite{mungse2023} provides a detailed examination of this
architecture, including the role of Virtual Routing and Forwarding
(VRF) tables in maintaining traffic isolation between different
customers. Each VRF on a PE router acts as a completely separate
routing table for a specific customer VPN. This isolation ensures
that routes and traffic from one customer are never visible to
another, even though both share the same physical infrastructure.

\subsubsection{Layer-2 MPLS VPNs}

Layer-2 MPLS VPNs present the service provider network as a virtual
switch to the customer. The customer sees nothing but Layer-2
circuits between their sites, and the provider handles all internal
routing and label switching. Two main variants exist. Virtual
Private Wire Service (VPWS) provides point-to-point Layer-2
circuits. Virtual Private LAN Service (VPLS) provides a
multi-point Layer-2 service that appears to the customer as a
single, shared LAN switch.

Fayed et al.~\cite{fayed2024} note that in a Layer-2 VPN, the
service provider network is transparent to the customer at Layer-3.
This means the customer is fully responsible for their own IP
addressing and routing, and point-to-point addresses must be
allocated at both customer sites from the same subnet. While this
simplicity gives customers full routing control, it also means that
the provider cannot optimize traffic paths based on Layer-3
information, which can limit efficiency in complex topologies.

\subsubsection{Layer-3 MPLS VPNs (BGP/MPLS VPNs)}

Layer-3 MPLS VPNs, defined in RFC~4364~\cite{rfc4364}, are the
most widely deployed MPLS VPN type in enterprise and carrier
environments. In an L3VPN, the service provider participates
directly in the customer's routing by maintaining separate VRF
instances for each customer on the PE routers. BGP with
Multiprotocol extensions (MP-BGP) distributes VPN routing
information between PE routers.

Two key concepts make L3VPNs function correctly. The first is the
Route Distinguisher (RD), a 64-bit value prepended to each customer
route that makes it globally unique, even if multiple customers use
overlapping IP address spaces. The second is the Route Target (RT),
a BGP community attribute that controls which routes are exported
from and imported into each VRF. By carefully configuring RTs,
service providers can implement complex connectivity models such as
full mesh, hub-and-spoke, and overlapping VPNs.

Mungse~\cite{mungse2023} implemented a complete L3VPN in GNS3 with
PE, P, and CE routers. The implementation demonstrated VRF-based
traffic isolation, BGP route distribution, and multi-site
reachability. Ping and traceroute tests confirmed that Site~A
traffic reached all designated Site~B destinations while remaining
completely isolated from other customers' traffic. Wireshark
captures confirmed BGP UPDATE messages carrying VPN route
information between PE routers.

Fayed et al.~\cite{fayed2024} extended this work to examine more
complex MPLS VPN scenarios. They implemented and validated three
advanced models on a live network emulation using GNS3 with Cisco
IOU software. The first was the simple VPN, where each customer's
sites communicate only among themselves. The second was the
overlapping VPN, where customer Headquarters sites communicate
with each other across VPN boundaries while branch sites remain
isolated. This is achieved by assigning additional Route Targets
to the HQ PE routers, enabling controlled route leaking between
otherwise separate VPNs. The third was the central services VPN,
where a shared server farm is accessible to all customers through
their individual VPNs while customers remain isolated from each
other. All three models were validated by inspecting BGP routing
tables of customer routers and confirming expected route presence
and absence.

The central services model is particularly relevant for enterprise
deployments, as it allows a shared security infrastructure, Active
Directory, or DNS service to serve multiple customer segments
without requiring those segments to communicate directly with
each other.

Huda and Andriyani~\cite{huda2025} further demonstrated how GRE
tunnels can be used in conjunction with OSPF multi-area designs,
showing that GRE and MPLS need not be viewed as competing
technologies but can complement each other in hybrid WAN
architectures.

% ---------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}

\subsection{Security Comparison}

The three technologies differ fundamentally in their approach to
security. IPSec provides the strongest built-in security, offering
encryption, integrity verification, and mutual authentication in
a single framework. GRE alone provides no security whatsoever.
MPLS VPNs rely on traffic isolation rather than encryption;
customer traffic is segregated at the routing layer through VRF
tables and label switching, but the traffic itself is not
encrypted in transit.

This distinction matters in practice. An MPLS VPN deployed over a
service provider's managed infrastructure is generally considered
secure because the provider controls the core network and prevents
unauthorized access. However, MPLS VPN traffic running over a
shared or untrusted backbone provides no protection for the data.
In such scenarios, IPSec must be added as an overlay to provide
encryption~\cite{fayed2024}. For organizations that must comply with
data protection regulations, IPSec or GRE-over-IPSec is the
appropriate choice, as it provides end-to-end encryption
regardless of the underlying transport.

\subsection{Scalability and Management}

MPLS VPNs have the strongest scalability story of the three
technologies. The VRF-based architecture separates customer routing
tables at PE routers, meaning P routers in the core carry no
VPN-specific state. Adding a new customer VPN requires configuration
only at the edge. This property allows service providers to support
thousands of enterprise customers on the same physical
infrastructure without increasing the complexity of the
core~\cite{mungse2023}.

GRE tunnels do not scale well in full-mesh deployments. If an
enterprise has $N$ sites and each must reach all others, $N(N-1)/2$
tunnel pairs must be configured. For even moderately sized networks
this becomes administratively burdensome. IPSec suffers from
similar scaling limitations, as each tunnel requires its own
Security Association (SA). MPLS also excels in supporting complex
connectivity models through the RT/RD mechanism, which allows
fine-grained control over site reachability --- including overlapping
and central services models --- with no equivalent in GRE or
IPSec~\cite{fayed2024}.

\subsection{Quality of Service}

MPLS supports QoS through the EXP bits in the label header, which
mark different traffic classes for differential treatment within
the provider network. This allows a service provider to guarantee
bandwidth, delay, and jitter for specific traffic classes, critical
for voice and video applications. Service Level Agreements for
delay-sensitive traffic are routinely backed by MPLS traffic
engineering in carrier networks~\cite{fayed2024}.

GRE and IPSec do not inherently support QoS. When traffic is
encapsulated in a GRE or IPSec tunnel, inner DSCP markings may or
may not be preserved in the outer header, depending on
configuration. In practice, QoS over GRE-IPSec tunnels requires
careful end-to-end planning and is much harder to guarantee than
in a managed MPLS environment.

\subsection{Cost and Deployment Context}

GRE and IPSec are software-based features available on virtually
all modern enterprise routers and firewalls, often at no
additional cost. They can be deployed quickly over any IP network,
including commodity Internet connections. For small to medium
enterprises with limited budgets, GRE-over-IPSec over broadband
Internet is a very cost-effective VPN solution.

MPLS VPNs typically require a service provider subscription.
However, for large enterprises with stringent SLA requirements,
the cost is often justified by the reliability, QoS guarantees,
and reduced management overhead. Fahriza et al.~\cite{fahriza2026}
demonstrated through detailed financial modeling that even a
self-managed GRE-IPSec deployment can be economically favorable,
showing a positive Net Present Value, an IRR of 84.80\%, and
a payback period of under one year for a three-branch Indonesian
enterprise.

\subsection{Summary Comparison}

Table~\ref{tab:comparison} summarizes the key differences across
the four configurations discussed in this paper.

\begin{table}[htbp]
\caption{Comparative Summary of Enterprise VPN Technologies}
\label{tab:comparison}
\centering
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lcccc}
\toprule
\textbf{Criterion} & \textbf{GRE} & \textbf{IPSec} & \textbf{GRE/IPSec} & \textbf{MPLS L3VPN} \\
\midrule
Encryption        & None    & Strong  & Strong  & None$^{*}$  \\
Authentication    & None    & Strong  & Strong  & Implicit    \\
Routing Support   & Full    & Limited & Full    & Full (BGP)  \\
QoS Support       & Basic   & Basic   & Basic   & Native      \\
Scalability       & Low     & Low     & Low     & High        \\
Configuration     & Low     & Medium  & Medium  & High        \\
Cost              & Low     & Low     & Low     & High        \\
Multicast         & Yes     & No      & Yes     & Yes         \\
Deployment Side   & Customer& Customer& Customer& Provider    \\
\bottomrule
\multicolumn{5}{l}{$^{*}$MPLS provides traffic isolation, not encryption.}
\end{tabular}
\end{table}

The table shows that no single technology dominates across all
criteria. GRE-over-IPSec is the most practical choice for
enterprises that manage their own WAN connectivity over the
Internet. MPLS L3VPN is the preferred choice for enterprises that
require carrier-grade QoS, scalability, and complex connectivity
models, and are willing to pay for a managed service. Organizations
with very high security requirements may choose to run IPSec over
their MPLS circuits as an additional protection layer against
insider threats at the service provider.

% ---------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

This paper reviewed three foundational enterprise VPN technologies
--- GRE, IPSec, and MPLS --- from the perspective of their technical
operation, security properties, and practical deployment
characteristics. The review covered tunnel-based approaches
including standalone GRE, IPSec in tunnel mode, and the widely
deployed combination of GRE-over-IPSec, as well as provider-managed
MPLS VPNs including both Layer-2 and Layer-3 variants with complex
deployment models.

Several conclusions emerge from this survey. First, GRE should
never be deployed alone in environments where data confidentiality
is required. Its value lies in routing flexibility, and it is best
used as an inner tunnel encapsulated within IPSec. Second, IPSec
remains the gold standard for encrypting VPN traffic, and IKEv2
with modern cipher suites such as AES-256 and SHA-256 should be
the mandatory configuration in any new deployment. Third, MPLS
L3VPNs offer capabilities in scalability, QoS, and flexible
connectivity that GRE and IPSec simply cannot match. For large
enterprise and service provider environments, MPLS with
BGP-based L3VPN remains the most powerful solution available.

Future work should examine the convergence of traditional VPN
technologies with Software-Defined Networking (SDN) and
cloud-native architectures. As enterprises increasingly move
workloads to the cloud, the boundaries between GRE, IPSec, MPLS,
and cloud-native VPN solutions are blurring, and new hybrid
architectures are emerging that combine elements of all three.

% ---------------------------------------------------------------
\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
